---
title: "GIS"
author: "Kauthar Ismail"
date: "2025-02-26"
output: html_document
---
We will be looking at the distribution of bees in the western cape  , 
with their vegetation types

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



### Load the necessary libraries 

```{r}
library(sf)
library(lwgeom)
library(tidyverse)
library(terra)
library(stars)
library(exactextractr)
library(htmltools)
library(cowplot)
library(hrbrthemes)
library(leaflet)
library(rosm)
library(ggspatial)
library(rnaturalearth)
library(mapview)
library(leafpop)
library(tmap)
library(rinat)
library(prettymapr)
library(wesanderson)

```

### Reading in the data
#### reading in 2024 Final Vegetation Map of South Africa,Lesotho and Swaziland (Shapefile)[Vector]

```{r}
veg<-st_read("Shapefile/NVM2024Beta_IEM5_11_01072024.shp")
```
### Check the coordinate reference systems

```{r}
st_crs(veg)
```
##### The vegetation data  uses WGS84 coordinate system with code =4326

### Looking at the data more closely 
```{r}
class(veg)
head(veg)
```

### Checking if the file exists 
```{r}
file.exists("Shapefile/NVM2024Beta_IEM5_11_01072024.shp")
```
### Subsetting the data to the Western Cape using coordinates
```{r}

# Define coordinates in WGS84 (longitude, latitude)
coords <- data.frame(
  id = 1:3,  # Example points
  lon = c(18, 18.5, 19.5),  # Longitudes
  lat = c(-33, -34, -34.5)  # Latitudes
)

# Convert to an sf object
coords_sf <- st_as_sf(coords, coords = c("lon", "lat"), crs = 4326)

# Print results
print(coords_sf)

# Extract transformed X, Y coordinates
transformed_coords <- st_coordinates(coords_sf)

# Combine with original data
coords_WGS84_df <- cbind(coords, transformed_coords)

print(coords_WGS84_df)
st_write(coords_sf, "lo19_coordinates.shp",append=FALSE)

```

### Transforming the data so that the CRS systems are the same and then cropping
```{r}
coords_sf<- st_transform(coords_sf, st_crs(veg))
veg <- st_crop(veg, coords_sf) 
st_crs(veg)
```

### Basic plotting looking at the different bioregions in the western cape 

```{r}
plot(veg[6],  main="Bioregions")

``` 


### Loading the iNaturalist data on Bee populations within the Western Cape

```{r}

#Call the data directly from iNat
pc <- get_inat_obs(taxon_name = "Bee",
                   bounds = c(-35, 18, -33.5, 19.5),
                   maxresults = 2000)

#View the first few rows of data
head(pc)
```
``` 
```

### Filter returned according to certain attributes with positional accuaracy less than 46,latitude less than 0, wild specimen and research quality 

```{r}
pc <- pc %>% filter(positional_accuracy<46 & 
                latitude<0 &
                !is.na(latitude) &
                captive_cultivated == "false" &
                quality_grade == "research")
```

### Checking the class of the data frame
```{r}
class(pc)
```

### Converting the data frame into a spatial object of class=sf

```{r}
#Make the dataframe a spatial object of class = "sf"
pc <- st_as_sf(pc, coords = c("longitude", "latitude"), crs = 4326)
#check the class
class(pc)

```
### Check the data frame now to ensure it is a sf and has "geometry"
```{r}
names(pc)
```

### Plotting graph of the bee distribution and adding basemats to the plots
```{r}

ggplot() + 
  annotation_map_tile(type = "osm", progress = "none") + 
  geom_sf(data=pc)+ggtitle("Bee distribution within the Western Cape") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

rosm::osm.types()

```
### Creating interactive maps with leaflet and mapview
```{r}
library(leaflet)
library(htmltools)
leaflet() %>%
  # Add default OpenStreetMap map tiles
  addTiles(group = "Default") %>%  
  # Add our points
  addCircleMarkers(data = pc,
                   group = "bees",
                   radius = 3, 
                   color = "yellow") 
```

### Creating a map showing the details of the common name and whether the specimen is from a garden and making a live link
```{r}
library(mapview)
library(leafpop)

lpc <- pc %>%
  mutate(click_url = paste("<b><a href='", url, "'>Link to iNat observation</a></b>"))
mapview(pc, 
        popup = 
          popupTable(lpc,
            zcol = c("user_login", "captive_cultivated","common_name", "click_url")))
```

### Get the remnants layer to reduce the number of cultivated records with the remaining extent of the vegetation types 

```{r}
vegr <- st_read("Shapefile/NVM2024Beta_IEM5_11_01072024.shp")
```

### Transform the data and intersecting it 
```{r}
pc1 <- st_transform(pc, st_crs(vegr)) 
dim(pc1) # 769 rows and 36 columns
intersect<- st_intersection(pc1, vegr)
dim(intersect) # 765 rows and 36 columns
```

### Plotting the graph of the intersections
```{r}
ggplot() + 
  annotation_map_tile(type = "osm", progress = "none") + 
  geom_sf(data=intersect)+labs(title=)+
  ggtitle("Map of the intersections within the remaining extent of natural vegetation") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 13, face = "bold"))
```


### Plot of the Bee distribution in the vegetation groups of the Western Cape
```{r}
library(wesanderson)
library(ggplot2)
library(wesanderson)

pal <- wes_palette("Darjeeling1", 28, type = "continuous")
ggplot() + 
  annotation_map_tile(type = "osm", progress = "none") + 
  geom_sf(data = intersect, aes(col = T_Name)) +
  scale_colour_manual(values = pal) +
  ggtitle("Distribution of the Bee population within the varying vegetation groups of the Western Cape") +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 9, face = "bold", margin = margin(t = 40, b = 60)),  
    plot.title.position = "plot",  # Ensures title positioning
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)  # Increase bottom margin
  ) +labs(
    fill = "Vegetation Type",   # Legend for 'fill'
    color = "Vegetation type")

```